Machine Learning Deriven Set of microRNAs as a Novel Biomarker for Myocardial Infarction Diagnosis

### 2023-06-02 Complete-V1/Complete-V1.Rmd
The Frist vesrion befor Dr. Kavousi Revision

---
title: "The First Draft"
authors:
  - name: "Mehrdad Samadishadlou"
    address: "Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran"
date: "`r Sys.Date()`"
link-citation: yes
bibliography: [For Article Refrence.bib, discussion.bib, Introduction.bib]
output:
  bookdown::word_document2: default
  bookdown::pdf_book:
#    base_format: rticles::bioinformatics_article
always_allow_html: true
---

@@

```{r, datasets, echo=FALSE}
df <- data.frame(Dataset = c("GSE59867", "GSE56609", "GSE54475"),
                 Platform = c("GPL6244", "GPL6244", "GPL6244"),
                 Healthy = c("-", 46, 5),
                 CAD = c(46, "-", "-"),
                 MI = c(111, "-", "-"),
                 Refrence = c("[@67]", "[@09]", "[@75]"))

@@

The PCA plots of the training samples are shown in Figure \@ref(Fig:PCA)A and B. As shown, healthy samples were separated from CAD or MI samples in primary data and also after conducting fRMA. In the RLE plot, there was a distinct difference between dataset means for all samples before conducting fRMA  (Figure \@ref(Fig:PCA)C). All datasets were rearranged around 0 in the RLE plot after conducting fRMA (Figure \@ref(Fig:PCA)D). Moreover, there was a clear change in inter-quantile distances, but the values still were over 0.1.

```{r PCA, echo=F, out.width="100%", Figure .cap="PCA and RLE plot for all samples before and after fRMA."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/PCA and RLE of trains.pdf")
```

@@

```{r , DEGstab, echo=F}
M_H <- read.table("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Fisher results/DEGs MI-Healthy.txt", check.names = F)

@@

According to the cutoff criterion of $FDR < 0.05$, there were `r dim(M_H)[1]` DEGs between the MI and the healthy samples. Among them, `r sum(M_H$logFC > 0)` were up-regulated, and `r sum(M_H$logFC < 0)` were down-regulated in MI compared to the healthy controls. In CAD and healthy groups comparison, we found `r dim(C_H)[1]` DEGs, of which `r sum(C_H$logFC > 0)` and and `r sum(C_H$logFC < 0)` DEGs were up- and `r down-regulated, respectively in CAD samples. In the MI and CAD groups, the number of DEGs was `r dim(M_C)[1]`, and the number of up- and down-regulated genes in MI samples were `r sum(M_C$logFC > 0)` and `r sum(M_C$logFC < 0)`, respectively in comparison with CAD samples. These data are summarized in Table \@ref(tab:DEGstab).

```{r venn, echo=F, out.width="60%", Figure .align='center', Figure .cap="Venn diagram for DEGs in CAD/Healthy, MI/Healthy, and MI/CAD comparison."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/venny.png")
```

The Venn diagram  in Figure \@ref(Fig:venn) shows that CAD and MI samples share a majority of their DEGs. From 860 DEGs of MI/healthy and 670 DEGs of CAD/healthy, 531 genes were common which is 62% of MI/healthy DEGs and 79% of CAD/healthy DEGs.

@@

To explore the biological classification of the DEGs, we performed GO and KEGG pathway enrichment analyses on MI-healthy and CAD-healthy DEGs. In MI versus healthy samples, GO enrichment analysis in the BP category, suggested that the DEGs were enriched in "immune response-regulating signaling pathway", "lymphocyte differentiation", "immune response-regulating cell surface receptor signaling pathway", and "leukocyte activation involved in immune response" (Figure \@ref(Fig:MIHEnrich)A). In the CC category, the DEGs were enriched in "secretory granule membrane", "azurophil granule", "ficolin-1-rich granule", "tertiary granule", and "ficolin-1-rich granule membrane" (Figure \@ref(Fig:MIHEnrich)B). In the MF category, the DEGs were involved in "cadherin binding" and "MHC class I protein binding" (Figure \@ref(Fig:MIHEnrich)C). KEGG pathway analysis indicated that the DEGs were related to the following pathways: "Chemokine signaling pathway", "Lipid and atherosclerosis", and "Hematopoietic cell lineage" (Figure \@ref(Fig:MIHEnrich)D).

```{r MIHEnrich, echo=F, out.width="95%", Figure .cap="Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the MI and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/enrich MI-Healthy.pdf")
```

The enrichment results for DEGs of CAD versus healthy samples were as follows. In the BP category, GO enrichment suggested that the DEGs were enriched in "positive regulation of defense response", "positive regulation of innate immune response", "mononuclear cell differentiation", and "positive regulation of response to external stimulus" (Figure \@ref(Fig:CADHEnrich)A). In the CC category, the DEGs were enriched in "azurophil granule", "ficolin-1-rich granule", and "ficolin-1-rich granule membrane" (Figure \@ref(Fig:CADHEnrich)B). In the MF category, the DEGs were involved in "lipoprotein particle receptor binding" and "NF-$\kappa$B binding" (Figure \@ref(Fig:CADHEnrich)C). KEGG pathway analysis indicated that the DEGs were related to the following pathways: "Chemokine signaling pathway", "Lipid and atherosclerosis", and "Hematopoietic cell lineage" (Figure \@ref(Fig:CADHEnrich)D).

```{r CADHEnrich, echo=F, out.width="95%", Figure .cap="Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the CAD and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/enrich CAD-Healthy.pdf")
```

@@

Among all DEGs, just miR-186, miR-32, and miR-21 were detected as differentially expressed miRNAs. The expression profile of these three miRNAs is presented in Figure \@ref(Fig:DEMexp). Additionally, The ROC curves of each miRNA for each layer are presented in Figure \@ref(Fig:miRROC). Using the logistic regression model the AUC for miR-21, miR-32, and miR-186 was 0.99, 1, and 0.91 respectively  (Figure \@ref(Fig:miRROC)A). Besides, the accuracy of each miRNA for classifying the samples into healthy and not-healthy groups on the test set was 0.92, 0.98, and 0.89 for miR-21, miR-32, and miR-186, respectively. Moreover, the ROC curve of each miRNA for classifying MI and CAD samples was presented in Figure \@ref(Fig:miRROC)B. For miR-21, miR-32, and miR-186, the AUC and accuracy on the test set were 0.85; 0.7; and 0.82, and  0.78; 0.67; and 0.74, respectively.

@@

```{r DEMexp, echo=F, out.width="50%", Figure .align='center', Figure .cap="Expression profile of miR-186, miR-21, and miR32 in Healthy, CAD, and MI samples."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/DEGs healthy not-healthy/DEMs Expression.pdf")
```

```{r miRROC, echo=F, out.width="95%", Figure .align='center', Figure .cap="ROC curve for single miRNAs on test set classification for (A) healthy and not-healthy samples and (B) CAD and MI samples."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/miRs ROCs.pdf")
```

@@

Although single miRNAs had acceptable performance, their predictive value could be improved even further by using them as a set. The ROC curve for the SVM model with an RBF kernel trained with all three miRNAs is presented in Figure \@ref(Fig:DEMROC)A. The model had a better performance in classification than single miRNAs.The AUC for the model is 1, and its accuracy on the test set was also 1. In Figure \@ref(Fig:DEMCF)A, the confusion matrix for the model is presented.

```{r DEMROC, echo=F, out.width="95%", Figure .align='center', Figure .cap="ROC curve for miRNAs in DEGs on test set classification for (A) healthy and not-healthy samples and (B) CAD and MI samples."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/DEMs ROCs h not h cad mi.pdf")
```

```{r DEMCF, echo=F, out.width="90%", Figure .align='center', Figure .cap="Confusion matrix on the test set for (A) An SVM model with RBF kernel for healthy and not-healthy and (B) An SVM model with linear kernel for CAD and MI samples classification."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/DEMs Conf. matrix.pdf")
```

@@

Different models were trained using expression values for three differentially miRNAs. The models' 10-fold cross-validated AUC and accuracy on the test set are reported in Figure \@ref(Fig:DEMmodels). The best model from both AUC and accuracy point-of-view was the SVM model with linear kernel. The AUC and accuracy for this model with its pre-set values were 0.93 and 0.82 respectively. The model was hyper-tuned for C and gamma hyper-parameters, and therefore the model showed better performance. The ROC curve of the hyper-tuned model is presented in Figure \@ref(Fig:DEMROC)B. For this model the AUC reached  0.95 and the accuracy improved to 0.85 (Table \@ref(tab:DEGsML)). Moreover, the sensitivity and specificity for the model on the test set were 0.91 and 0.71 respectively. The confusion matrix for the hyper-tuned model is illustrated in Figure \@ref(Fig:DEMCF)B.

```{r DEMmodels, echo=F, out.width="65%", Figure .align='center', Figure .cap="Area under the curve (AUC) and accuracy of different models trained with three miRNAs in DEGs on the test set."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/Models DEGs mirs.pdf")
```

@@

After calculating the AUC for each miRNA for the classification of MI and CAD samples, miRNAs were sorted, and their performance was investigated as a set. The metric of choice for selecting the best set was AUC. As shown in Figure \@ref(Fig:AUCset), the AUC increased until the number of miRNAs in the set reached six, and after that, it dropped. The AUC for separating MI samples from CAD using these miRNAs was 0.93. The miRNAs in the set were miR-29a, miR-197, miR-142, miR-21, miR-155, and miR-320C1. The expression values of these miRNAs in healthy, CAD, and MI samples are presented in Figure \@ref(Fig:AUCexp). The ROC curve of the selected miRNAs for MI and CAD sample classification is illustrated in Figure \@ref(Fig:miRROC)B.

```{r AUCset, echo=F, out.width="65%", Figure .align='center', Figure .cap="Area under the curve (AUC) for sets containing an increasing number of miRNAs with the highest individual AUC in MI/CAD separation."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/AUC for cad mi mirs.pdf")
```

```{r AUCexp, echo=F, out.width="90%", Figure .align='center', Figure .cap="Expression profile of has-miR-29A, has-miR-197, has-miR-142, has-miR-21, has-miR-155, and has-miR-320C1 in Healthy, CAD, and MI samples."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/ExpressionforAUC.pdf")
```

@@

Using the selected set, an SVM model with an RBF kernel was trained to separate healthy from not-healthy samples. The ROC curve for the model is presented in Figure \@ref(Fig:AUCROC)A and the confusion matrix is illustrated in Figure \@ref(Fig:AUCCF)A. Both AUC and accuracy for the model on the test set were equal to 1.

```{r AUCROC, echo=F, out.width="95%", Figure .align='center', Figure .cap="ROC curve for the set of miRNAs selected by AUC on test set classification. (A) SVM with RBF kernel for healthy and not-healthy samples classification. (B) Logistic regression model for CAD and MI samples classification. (C) SVM with polynomial kernel for CAD and MI samples classification. "}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/AUC best model ROCs.pdf")
```

```{r AUCCF, echo=F, out.width="90%", Figure .align='center', Figure .cap="Confusion matrix on the test set for (A) SVM with RBF kernel for healthy and not-healthy samples classification. (B) Logistic regression model for CAD and MI samples classification. (C) SVM with polynomial kernel for CAD and MI samples classification."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/AUC Conf. matrix.pdf")
```

@@

To find the best model for training the best set, different models were trained using their pre-set values. Their AUC and accuracy results on the test set are presented in Figure \@ref(Fig:AUCmodels). The best model from the AUC point-of-view was the LR and from the accuracy point-of-view, it was the SVM model with a polynomial kernel. For the LR model the AUC and accuracy were 0.92 and 0.81, respectively; and for the SVM model with a polynomial kernel, the values were 0.91 and 0.84, respectively. Both models were hyper-tuned and the ROC curve for their best performance is presented in Figure \@ref(Fig:AUCROC)B and C. The AUC and accuracy for the LR model increased to 0.94 and 0.88, respectively. For the SVM model with a polynomial kernel, these values increased to 0.95 and 0.88, respectively (Table \@ref(tab:AUCML)). The sensitivity for the LR and SVM models were 1 and 0.97, respectively; and the specificity for them was 0.57 and 0.64, respectively. The confusion matrix for both models is illustrated in Figure \@ref(Fig:AUCCF)B and C.

```{r AUCmodels, echo=F, out.width="65%", Figure .align='center', Figure .cap="Area under the curve (AUC) and accuracy of different models trained with three miRNAs in DEGs on the test set."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/Models AUC mirs.pdf")
```

@@

In this study, we combined three GEO datasets of healthy, CAD, and MI samples. Having these samples set alongside bioinformatics analysis and ML means, it is possible to identify potential biomarker sets and also effective therapeutic targets. The results of the DEG analysis (Table \@ref(tab:DEGstab) and Figure \@ref(Fig:venn)) are proof of the close relationship between the MI and CAD samples. Interestingly, functional enrichment analysis demonstrated that DEGs in both healthy/CAD and healthy/MI were strongly correlated to immune cell response which is a major cellular part of PBMCs. Here, two different sets of miRNAs were used as biomarker sets for sample classification. miR-21; miR-32; and miR-186 were selected as differentially expressed miRNAs, and miR-21; miR-29a; miR-142; miR155; miR-197; and miR-320c1 were selected according to their AUC values. As shown in Figure \@ref(Fig:miRROC), all other miRNAs selected with both approaches had AUC-ROC over 0.9 for the isolation of healthy and not-healthy samples except for miR-142 and miR-29a. Data confirmed that the real challenge is to classify CAD and MI samples because of close overlap. Of 8 miRNAs under investigation in both approaches except for miR-32, all miRNAs had an AUC-ROC over 0.8 for the discrimination of CAD and MI samples. Besides, the high AUC-ROC values of miRNAs confirms their high potential as biomarkers.

@@

@@

>>>>

---
title: Machine Learning Deriven Set of microRNAs as a Novel Biomarker for Myocardial Infarction Diagnosis
#titlerunning: Short form of title (if too long for head)
authorrunning: M. Samadishadlou et al.
#thanks: | 
#    Grants or other notes about the article that should go on the front 
#    page should be placed here. General acknowledgments should be placed at the
#    end of the article.

authors: 
- name: Mehrdad Samadishadlou
  address: Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran
  email: samadishadlou@tbzmed.ac.ir

- name: Reza Rahbarghazi
  address: Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran
  email: rahbarghazir@tbzmed.ac.ir

- name: Farhad Bani
  address: Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran
  email: bainf@tbzmed.ac.ir

keywords:
- microRNA
- Machine Learning
- Myocardial Infarction

#PACS: 
#- PAC1
#- superPAC

#MSC:
#- MSC code 1
#- MSC code 2

always_allow_html: true

abstract: |
  MicroRNAs (miRNAs) play a crucial role in regulating adaptive and maladaptive responses in cardiovascular diseases, making them attractive targets for potential therapeutics. However, their potential as novel biomarkers for diagnosing cardiovascular diseases needs to be evaluated systematically. In this study, we aimed to identify a key set of miRNA biomarkers using integrated bioinformatics and machine learning analysis. We combined and analyzed three gene expression datasets from the Gene Expression Omnibus (GEO) database, which contained peripheral blood mononuclear cells samples from individuals with myocardial infarction (MI), stable coronary artery disease (CAD), and healthy. Additionally, we selected a set of miRNAs based on their area under the curve of the receiver operating characteristic (AUC-ROC) for separating CAD/MI samples. We designed a two-layer architecture for sample classification, with the first layer isolating healthy samples from not-healthy ones, and the second layer classifying stable CAD and MI samples. We trained different machine learning models using both biomarker sets and evaluated their performance on a test set. We identified miR-21, miR-186, and miR-32 as the only miRNAs in the differentially expressed genes, and a set included miR-21, miR-155, miR-142, miR-197, miR-29A, and miR-320C1 as the optimum set of miRNA selected by their AUC-ROC. Both biomarker sets were able to distinguish healthy from not-healthy samples with complete accuracy. The best performance for classification of CAD and MI was achieved with an SVM model trained using the biomarker set selected by AUC-ROC, with an AUC of 0.95 and an accuracy of 0.88. Our study demonstrates that miRNA signatures derived from peripheral blood could serve as valuable novel biomarkers for cardiovascular diseases.

output:
  rticles::springer_article

bibliography: [ForArticleRefrence.bib, discussion.bib, Introduction.bib]
biblio-style: spmpsci
# bibstyle options spbasic(default), spphys, spmpsci
---

@@

```{r datasets, echo=FALSE}
df <- data.frame(Dataset = c("GSE59867", "GSE56609", "GSE54475"),
                 Platform = c("GPL6244", "GPL6244", "GPL6244"),
                 Healthy = c("-", 46, 5),
                 CAD = c(46, "-", "-"),
                 MI = c(111, "-", "-"),
                 Refrence = c("[@67]", "[@09]", "[@75]"))

@@

The PCA plots of the training samples are shown in Figure \@ref(fig:PCA)A and B. As shown, healthy samples were separated from CAD or MI samples in primary data and also after conducting fRMA. In the RLE plot, there was a distinct difference between dataset means for all samples before conducting fRMA (Figure \@ref(fig:PCA)C). All datasets were rearranged around 0 in the RLE plot after conducting fRMA (Figure \@ref(fig:PCA)D). Moreover, there was a clear change in inter-quantile distances, but the values still were over 0.1.

```{r PCA, echo=F, out.width="100%", fig.cap="PCA and RLE plot for all samples before and after fRMA."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/PCA and RLE of trains.pdf")
```

@@

```{r DEGstab, echo=F}
M_H <- read.table("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Fisher results/DEGs MI-Healthy.txt", check.names = F)

@@

According to the cutoff criterion of $FDR < 0.05$, there were `r dim(M_H)[1]` DEGs between the MI and the healthy samples. Among them, `r sum(M_H$logFC > 0)` were up-regulated, and `r sum(M_H$logFC < 0)` were down-regulated in MI compared to the healthy controls. In CAD and healthy groups comparison, we found `r dim(C_H)[1]` DEGs, of which `r sum(C_H$logFC > 0)` and and `r sum(C_H$logFC < 0)` DEGs were up- and down-regulated, respectively in CAD samples. In the MI and CAD groups, the number of DEGs was `r dim(M_C)[1]`, and the number of up- and down-regulated genes in MI samples were `r sum(M_C$logFC > 0)` and `r sum(M_C$logFC < 0)`, respectively in comparison with CAD samples. These data are summarized in Table \@ref(tab:DEGstab).

```{r venn, echo=F, out.width="60%", fig.align='center', fig.cap="Venn diagram for DEGs in CAD/Healthy, MI/Healthy, and MI/CAD comparison."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/venny.png")
```

The Venn diagram in Figure \@ref(fig:venn) shows that CAD and MI samples share a majority of their DEGs. From 860 DEGs of MI/healthy and 670 DEGs of CAD/healthy, 531 genes were common which is 62% of MI/healthy DEGs and 79% of CAD/healthy DEGs.

@@

To explore the biological classification of the DEGs, we performed GO and KEGG pathway enrichment analyses on MI-healthy and CAD-healthy DEGs. In MI versus healthy samples, GO enrichment analysis in the BP category, suggested that the DEGs were enriched in "immune response-regulating signaling pathway", "lymphocyte differentiation", "immune response-regulating cell surface receptor signaling pathway", and "leukocyte activation involved in immune response" (Figure \@ref(fig:MIHEnrich)A). In the CC category, the DEGs were enriched in "secretory granule membrane", "azurophil granule", "ficolin-1-rich granule", "tertiary granule", and "ficolin-1-rich granule membrane" (Figure \@ref(fig:MIHEnrich)B). In the MF category, the DEGs were involved in "cadherin binding" and "MHC class I protein binding" (Figure \@ref(fig:MIHEnrich)C). KEGG pathway analysis indicated that the DEGs were related to the following pathways: "Chemokine signaling pathway", "Lipid and atherosclerosis", and "Hematopoietic cell lineage" (Figure \@ref(fig:MIHEnrich)D).

```{r MIHEnrich, echo=F, out.width="95%", fig.cap="Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the MI and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/enrich MI-Healthy.pdf")
```

The enrichment results for DEGs of CAD versus healthy samples were as follows. In the BP category, GO enrichment suggested that the DEGs were enriched in "positive regulation of defense response", "positive regulation of innate immune response", "mononuclear cell differentiation", and "positive regulation of response to external stimulus" (Figure \@ref(fig:CADHEnrich)A). In the CC category, the DEGs were enriched in "azurophil granule", "ficolin-1-rich granule", and "ficolin-1-rich granule membrane" (Figure \@ref(fig:CADHEnrich)B). In the MF category, the DEGs were involved in "lipoprotein particle receptor binding" and "NF-$\kappa$B binding" (Figure \@ref(fig:CADHEnrich)C). KEGG pathway analysis indicated that the DEGs were related to the following pathways: "Chemokine signaling pathway", "Lipid and atherosclerosis", and "Hematopoietic cell lineage" (Figure \@ref(fig:CADHEnrich)D).

```{r CADHEnrich, echo=F, out.width="95%", fig.cap="Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the CAD and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/enrich CAD-Healthy.pdf")
```

@@

Among all DEGs, just miR-186, miR-32, and miR-21 were detected as differentially expressed miRNAs. The expression profile of these three miRNAs is presented in Figure \@ref(fig:DEMexp). Additionally, The ROC curves of each miRNA for each layer are presented in Figure \@ref(fig:miRROC). Using the logistic regression model the AUC for miR-21, miR-32, and miR-186 was 0.99, 1, and 0.91 respectively (Figure \@ref(fig:miRROC)A). Besides, the accuracy of each miRNA for classifying the samples into healthy and not-healthy groups on the test set was 0.92, 0.98, and 0.89 for miR-21, miR-32, and miR-186, respectively. Moreover, the ROC curve of each miRNA for classifying MI and CAD samples was presented in Figure \@ref(fig:miRROC)B. For miR-21, miR-32, and miR-186, the AUC and accuracy on the test set were 0.85; 0.7; and 0.82, and 0.78; 0.67; and 0.74, respectively.

@@

```{r DEMexp, echo=F, out.width="50%", fig.align='center', fig.cap="Expression profile of miR-186, miR-21, and miR32 in Healthy, CAD, and MI samples."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/DEGs healthy not-healthy/DEMs Expression.pdf")
```

```{r miRROC, echo=F, out.width="95%", fig.align='center', fig.cap="ROC curve for single miRNAs on test set classification for (A) healthy and not-healthy samples and (B) CAD and MI samples."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/miRs ROCs.pdf")
```

@@

Although single miRNAs had acceptable performance, their predictive value could be improved even further by using them as a set. The ROC curve for the SVM model with an RBF kernel trained with all three miRNAs is presented in Figure \@ref(fig:DEMROC)A. The model had a better performance in classification than single miRNAs.The AUC for the model is 1, and its accuracy on the test set was also 1. In Figure \@ref(fig:DEMCF)A, the confusion matrix for the model is presented.

```{r DEMROC, echo=F, out.width="95%", fig.align='center', fig.cap="ROC curve for miRNAs in DEGs on test set classification for (A) healthy and not-healthy samples and (B) CAD and MI samples."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/DEMs ROCs h not h cad mi.pdf")
```

```{r DEMCF, echo=F, out.width="90%", fig.align='center', fig.cap="Confusion matrix on the test set for (A) An SVM model with RBF kernel for healthy and not-healthy and (B) An SVM model with linear kernel for CAD and MI samples classification."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/DEMs Conf. matrix.pdf")
```

@@

Different models were trained using expression values for three differentially miRNAs. The models' 10-fold cross-validated AUC and accuracy on the test set are reported in Figure \@ref(fig:DEMmodels). The best model from both AUC and accuracy point-of-view was the SVM model with linear kernel. The AUC and accuracy for this model with its pre-set values were 0.93 and 0.82 respectively. The model was hyper-tuned for C and gamma hyper-parameters, and therefore the model showed better performance. The ROC curve of the hyper-tuned model is presented in Figure \@ref(fig:DEMROC)B. For this model the AUC reached 0.95 and the accuracy improved to 0.85 (Table \@ref(tab:DEGsML)). Moreover, the sensitivity and specificity for the model on the test set were 0.91 and 0.71 respectively. The confusion matrix for the hyper-tuned model is illustrated in Figure \@ref(fig:DEMCF)B.

```{r DEMmodels, echo=F, out.width="65%", fig.align='center', fig.cap="Area under the curve (AUC) and accuracy of different models trained with three miRNAs in DEGs on the test set."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/Models DEGs mirs.pdf")
```

@@

After calculating the AUC for each miRNA for the classification of MI and CAD samples, miRNAs were sorted, and their performance was investigated as a set. The metric of choice for selecting the best set was AUC. As shown in Figure \@ref(fig:AUCset), the AUC increased until the number of miRNAs in the set reached six, and after that, it dropped. The AUC for separating MI samples from CAD using these miRNAs was 0.93. The miRNAs in the set were miR-29a, miR-197, miR-142, miR-21, miR-155, and miR-320C1. The expression values of these miRNAs in healthy, CAD, and MI samples are presented in Figure \@ref(fig:AUCexp). The ROC curve of the selected miRNAs for MI and CAD sample classification is illustrated in Figure \@ref(fig:miRROC)B.

```{r AUCset, echo=F, out.width="65%", fig.align='center', fig.cap="Area under the curve (AUC) for sets containing an increasing number of miRNAs with the highest individual AUC in MI/CAD separation."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/AUC for cad mi mirs.pdf")
```

```{r AUCexp, echo=F, out.width="90%", fig.align='center', fig.cap="Expression profile of has-miR-29A, has-miR-197, has-miR-142, has-miR-21, has-miR-155, and has-miR-320C1 in Healthy, CAD, and MI samples."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/ExpressionforAUC.pdf")
```

@@

Using the selected set, an SVM model with an RBF kernel was trained to separate healthy from not-healthy samples. The ROC curve for the model is presented in Figure \@ref(fig:AUCROC)A and the confusion matrix is illustrated in Figure \@ref(fig:AUCCF)A. Both AUC and accuracy for the model on the test set were equal to 1.

```{r AUCROC, echo=F, out.width="95%", fig.align='center', fig.cap="ROC curve for the set of miRNAs selected by AUC on test set classification. (A) SVM with RBF kernel for healthy and not-healthy samples classification. (B) Logistic regression model for CAD and MI samples classification. (C) SVM with polynomial kernel for CAD and MI samples classification. "}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/AUC best model ROCs.pdf")
```

```{r AUCCF, echo=F, out.width="90%", fig.align='center', fig.cap="Confusion matrix on the test set for (A) SVM with RBF kernel for healthy and not-healthy samples classification. (B) Logistic regression model for CAD and MI samples classification. (C) SVM with polynomial kernel for CAD and MI samples classification."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/AUC Conf. matrix.pdf")
```

@@

To find the best model for training the best set, different models were trained using their pre-set values. Their AUC and accuracy results on the test set are presented in Figure \@ref(fig:AUCmodels). The best model from the AUC point-of-view was the LR and from the accuracy point-of-view, it was the SVM model with a polynomial kernel. For the LR model the AUC and accuracy were 0.92 and 0.81, respectively; and for the SVM model with a polynomial kernel, the values were 0.91 and 0.84, respectively. Both models were hyper-tuned and the ROC curve for their best performance is presented in Figure \@ref(fig:AUCROC)B and C. The AUC and accuracy for the LR model increased to 0.94 and 0.88, respectively. For the SVM model with a polynomial kernel, these values increased to 0.95 and 0.88, respectively (Table \@ref(tab:AUCML)). The sensitivity for the LR and SVM models were 1 and 0.97, respectively; and the specificity for them was 0.57 and 0.64, respectively. The confusion matrix for both models is illustrated in Figure \@ref(fig:AUCCF)B and C.

```{r AUCmodels, echo=F, out.width="65%", fig.align='center', fig.cap="Area under the curve (AUC) and accuracy of different models trained with three miRNAs in DEGs on the test set."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/Models AUC mirs.pdf")
```

@@

In this study, we combined three GEO datasets of healthy, CAD, and MI samples. Having these samples set alongside bioinformatics analysis and ML means, it is possible to identify potential biomarker sets and also effective therapeutic targets. The results of the DEG analysis (Table \@ref(tab:DEGstab) and Figure \@ref(fig:venn)) are proof of the close relationship between the MI and CAD samples. Interestingly, functional enrichment analysis demonstrated that DEGs in both healthy/CAD and healthy/MI were strongly correlated to immune cell response which is a major cellular part of PBMCs. Here, two different sets of miRNAs were used as biomarker sets for sample classification. miR-21; miR-32; and miR-186 were selected as differentially expressed miRNAs, and miR-21; miR-29a; miR-142; miR155; miR-197; and miR-320c1 were selected according to their AUC values. As shown in Figure \@ref(fig:miRROC), all other miRNAs selected with both approaches had AUC-ROC over 0.9 for the isolation of healthy and not-healthy samples except for miR-142 and miR-29a. Data confirmed that the real challenge is to classify CAD and MI samples because of close overlap. Of 8 miRNAs under investigation in both approaches except for miR-32, all miRNAs had an AUC-ROC over 0.8 for the discrimination of CAD and MI samples. Besides, the high AUC-ROC values of miRNAs confirms their high potential as biomarkers.

@@

# Conclusion

In summary, we derived a set of miRNA biomarkers by comparing MI samples to both healthy and CAD samples. We found that the SVM model performed best in both the first layer, which separated healthy and not-healthy samples, and the second layer, which classified MI/CAD samples. The set of miRNAs selected based on their AUC values had slightly better performance in the second layer. Overall, our two-layer structure achieved a weighted accuracy of 0.91. This demonstrates the potential for combining bioinformatics and machine learning techniques to identify novel biomarkers and gain a deeper understanding of myocardial infarction.

@@

\begin{align}
a^2+b^2=c^2
\end{align}
